---
title: "Wines-Prediction"
author: "Hayoung Cheon"
date: "2024-12-15"
output: html_document
---

```{r}
library(dplyr)
library(MASS)
library(car)
library(ggplot2)
library(tidyverse)
library(GGally)
library(plotly)
library(olsrr)
library(caret)
library(leaps)
library(corrplot)
```
## Load Data
```{r}
wines <- read.csv("/Users/oyunm/Desktop/SMU/DS-6306-Doing-the-Data-Science/Project 2/Wine Train Set.csv")
head(wines)

wines$location <- gsub("Califormia", "California", wines$location, ignore.case=TRUE)
table(wines$quality, wines$location)
```



# base model
According to p-values citric.acid is not significant variable to the model.
```{r}
head(wines)

# drop ID & quality
wine_filtered <-wines%>%select(-ID)

wine_filtered$quality <- as.numeric(as.character(wine_filtered$quality))

fit <- lm(quality ~ ., data=wine_filtered  )
summary(fit)
```

# Correlation Matrix

According to Correlation MAtrix the variables that are highly correlated to the quality was volatile.acidity, chlorides, density and alcohol. We 
```{r}
excluded_columns <- c("ID","type", "location")
selected_data <- wines[, setdiff(names(wines), excluded_columns)]

str(selected_data)
selected_data$quality <- as.numeric(selected_data$quality)

# Compute correlation matrix for the selected columns
cor_matrix <- cor(selected_data)
print(cor_matrix)

# Visualize correlation matrix

# Install and load the corrplot package
#install.packages("corrplot")
library(corrplot)

# Plot the correlation matrix
corrplot(
  cor_matrix, 
  method = "circle",
  addCoef.col = "black",
  number.cex = 0.8,          # Adjust text size for values
  tl.col = "black",          # Black color for labels
  tl.cex = 0.8  
  )
```

# Check Multicollineaarity with Variance Inflation Factor(VIF)

Density shows the highest VIF, and according to correlation matrix it has big correlation to all other variables that have high correlation with quality.
```{r}

# Ensure they are factor
wine_filtered$type <- as.factor(wine_filtered$type)
wine_filtered$location <- as.factor(wine_filtered$location)

# fit a model without citric.acid
fit <- lm(quality ~ fixed.acidity + volatile.acidity +residual.sugar+chlorides + free.sulfur.dioxide +total.sulfur.dioxide + density +pH + sulphates +alcohol + type + location, data=wine_filtered)

# 
summary(fit)
vif(fit)
AIC(fit)

```
## Without density
```{r}
# fit a model without citric.acid
fit <- lm(quality ~ volatile.acidity+chlorides + free.sulfur.dioxide +total.sulfur.dioxide + density+ I(density^2) + sulphates +alcohol + I(alcohol^2) + type + location, data=wine_filtered)

# 
summary(fit)
vif(fit)
AIC(fit)
```

# Check residual plots for assumption met
```{r}

par(mfrow = c(2, 2))  # Arrange plots in a 2x2 grid
plot(fit)

```
# Explore influential points found during EDA
```{r}
wine_filtered[5414,]
wine_filtered[4326,]

# 5151,3152 : Restrict Range of density outliers to 1.05. 
subset_wines <- wines %>% filter(density > 1.01)
print(subset_wines)

# 1056 : Volatile.acidity Morethan 1.4 which is legal limit
subset_wines<- wines %>% filter(volatile.acidity > 1.5)
print(subset_wines)

wine_filt2 <- wine_filtered[-c(5151,3152, 1056, 5414,4326 ),]
fit2 <- lm(quality ~ fixed.acidity + volatile.acidity+chlorides + free.sulfur.dioxide +total.sulfur.dioxide + density +pH + sulphates +alcohol + type + location, data=wine_filt2)
```
# Check high leverage rows
```{r}
# Calculate leverage values
leverage_values <- hatvalues(fit2)

# Set high-leverage threshold
n <- nrow(wine_filt2)
p <- length(coef(fit2))
threshold <- 2 * (p + 1) / n

# Identify high-leverage rows
high_leverage_rows <- which(leverage_values > threshold)

# View high-leverage points
print(high_leverage_rows)
print(wine_filt2[high_leverage_rows, ])

```
```{r}
residuals <- resid(fit)

ggplot(data = NULL, aes(x = residuals)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") +
  labs(title = "Residuals Distri
       bution", x = "Residuals", y = "Frequency")


```

# Quadratic
```{r}
fit_quad <- lm(quality ~ .  + I(volatile.acidity^2) +
                 I(residual.sugar^2)  +
                 I(density^2) + I(pH^2) + I(alcohol^2) - citric.acid - chlorides, 
               data = wine_filt2)
summary(fit_quad)
AIC(fit_quad)
vif(fit_quad)

fit2 <-lm(quality ~ .  + I(volatile.acidity^2) +
                 I(residual.sugar^2)  +
                 I(density^2) + I(pH^2) + I(alcohol^2) - citric.acid - chlorides, 
               data = wine_filt2)
```

# Train DAta and find Mean Absolute Error
```{r}
set.seed(123)
train_idx <- sample(seq_len(nrow(wine_filt2)), size = 0.7 * nrow(wine_filt2))
train_data <- wine_filt2[train_idx, ]
test_data <- wine_filt2[-train_idx, ]

# Fit the regression model


fit2 <-lm(quality ~ fixed.acidity + volatile.acidity+chlorides + free.sulfur.dioxide +total.sulfur.dioxide + sulphates +alcohol + type + location  + I(volatile.acidity^2) +I(residual.sugar^2)  +I(density^2)  + I(alcohol^2), data = wine_filt2)
vif(fit2)


# Predict on the test data
predictions <- predict(fit2, test_data)

# Evaluate performance using Mean Absolute Error (MAE)
actual <- test_data$quality
mae <- mean(abs(predictions - actual))  # Calculate MAE
mae

```

# Train and test data with cubic model
```{r}
set.seed(123)
train_idx <- sample(seq_len(nrow(wine_filt2)), size = 0.7 * nrow(wine_filt2))
train_data <- wine_filt2[train_idx, ]
test_data <- wine_filt2[-train_idx, ]

# Fit the regression model


fit3 <-lm(log(quality) ~  volatile.acidity + log(free.sulfur.dioxide)  + log(sulphates) +alcohol + type + location  + I(volatile.acidity^2) +I(log(residual.sugar)^2)  +I(log(density)^2)  + I(alcohol^2)+ I(volatile.acidity^3)+ I(citric.acid^3)+ I(log(chlorides)^3) + I(total.sulfur.dioxide^3) + I(alcohol^3) , data = wine_filt2)
vif(fit3)
summary(fit3)


# Predict on the test data
predictions <- exp(predict(fit3, test_data))

# Evaluate performance using Mean Absolute Error (MAE)
actual <- test_data$quality
mae <- mean(abs(predictions - actual))  # Calculate MAE
mae
```

# Again check for Outliers and remove(restrict range)
```{r}
par(mfrow = c(2, 2))
plot(fit3)

cooksd <- cooks.distance(fit3)
outliers <- which(cooksd > (4 / nrow(wine_filt2)))  # Common threshold
wine_filt2_cleaned <- wine_filt2[-outliers, ]


```

## Train and test data with cubic model
After restricting the range of outliers, the mae improved
```{r}
set.seed(123)
train_idx <- sample(seq_len(nrow(wine_filt2_cleaned)), size = 0.7 * nrow(wine_filt2_cleaned))
train_data <- wine_filt2_cleaned[train_idx, ]
test_data <- wine_filt2_cleaned[-train_idx, ]

# Fit the regression model


fit4 <-lm(log(quality) ~  volatile.acidity + log(free.sulfur.dioxide)  + log(sulphates) +alcohol  + location  + I(volatile.acidity^2) +I(log(residual.sugar)^2)  +I(log(density)^2)  + I(alcohol^2)+ I(volatile.acidity^3)+ I(citric.acid^3)+ I(log(chlorides)^3) + I(total.sulfur.dioxide^3) + I(alcohol^3) , data = wine_filt2_cleaned)
vif(fit4)
summary(fit4)


# Predict on the test data
predictions <- exp(predict(fit4, test_data))


# Evaluate performance using Mean Absolute Error (MAE)
actual <- test_data$quality
mae <- mean(abs(predictions - actual))  # Calculate MAE
mae
```


## Cross-validation to validate model's performance

```{r}
train_control <- trainControl(method = "cv", number = 5)
fit_cv <- train(log(quality) ~  volatile.acidity + log(free.sulfur.dioxide)  + log(sulphates) +alcohol  + location  + I(volatile.acidity^2) +I(log(residual.sugar)^2)  +I(log(density)^2)  + I(alcohol^2)+ I(volatile.acidity^3)+ I(citric.acid^3)+ I(log(chlorides)^3) + I(total.sulfur.dioxide^3) + I(alcohol^3) , data = wine_filt2_cleaned, method = "lm", trControl = train_control)
fit_cv
```

# Conclusion
By using Regression model, the best model fit was log(quality) ~  volatile.acidity + log(free.sulfur.dioxide)  + log(sulphates) +alcohol + location  + I(volatile.acidity^2) +I(log(residual.sugar)^2)  +I(log(density)^2)  + I(alcohol^2)+ I(volatile.acidity^3)+ I(citric.acid^3)+ I(log(chlorides)^3) + I(total.sulfur.dioxide^3) + I(alcohol^3). Although since it is polynomial, we have to look out for overfitting. 

# Create CSV file with the prediction
```{r}
data_test <- read.csv("/Users/oyunm/Desktop/SMU/DS-6306-Doing-the-Data-Science/Project 2/Wine Test Set.csv")

data_loc_type <- read.csv("/Users/oyunm/Desktop/SMU/DS-6306-Doing-the-Data-Science/Project 2/Wine Types And Locations.csv")
subset_data <- data_loc_type[5464:nrow(data_loc_type), ]

# Merge data_test + subset_data(this has type and location)
joined_data_test <- left_join(data_test, subset_data, by = "ID")

# In Joined_data_test taking out any missing data from type.
joined_data_test %>% filter(!is.na(joined_data_test$type))

#There will be missing data after prediction since we are omitting type that are missing. for those data we can set prediction with median number.
joined_data_test$quality <- 6

fit4 <-lm(log(quality) ~  volatile.acidity + log(free.sulfur.dioxide)  + log(sulphates) +alcohol + type + location  + I(volatile.acidity^2) +I(log(residual.sugar)^2)  +I(log(density)^2)  + I(alcohol^2)+ I(volatile.acidity^3)+ I(citric.acid^3)+ I(log(chlorides)^3) + I(total.sulfur.dioxide^3) + I(alcohol^3) , data = wine_filt2_cleaned)
vif(fit4)
summary(fit4)

wine_filt2$type <- as.factor(wine_filt2$type)
joined_data_test$type <- as.factor(joined_data_test$type)

# type's level mining
# Replace "Red" with "red" in the location column
joined_data_test$type <- gsub("Red", "red", joined_data_test$type, ignore.case = TRUE)

#Replace "White" with "white" in the type column
joined_data_test$type <- gsub("White", "white", joined_data_test$type, ignore.case = TRUE)

joined_data_test$location <- gsub("Califormia", "California", joined_data_test$location, ignore.case = TRUE)


#Taking care of missing data
#joined-data-test <- na.omit(joined_data_test)

# Levels in the training data
levels(wine_filt2$type)

# Levels in the test data
levels(joined_data_test$type)

# Ensure the type column is a factor
joined_data_test$type <- as.factor(joined_data_test$type)

# Ensure type is a factor in both data sets
wine_filt2$type <- as.factor(wine_filt2$type)
joined_data_test$type <- as.factor(joined_data_test$type)

# Match the levels of joined_data_test$type with wine_filt2$type
joined_data_test$type <- factor(joined_data_test$type, levels = levels(wine_filt2$type))


# Match levels with the training data
levels(joined_data_test$type) <- levels(wine_filt2$type)


# Predict on the test data
predictions <- exp(predict(fit4, newdata=joined_data_test))
joined_data_test$quality <- predictions


output <- joined_data_test |> dplyr::select("ID", "quality")

# Write Results
write.csv(output, file = "wine Prediction Set final.csv", row.names = FALSE)
```